---
title: "Comparison to MultiAssayExperiment"
author: "Yuchen Li"
date: "Revised: 17 June, 2025"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{MultiAssayExperiment-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# Introduction

The `FlexAssays` is conceptually related to [`MultiAssayExperiment`](https://www.bioconductor.org/packages/release/bioc/html/MultiAssayExperiment.html) (MAE), as both aim to store and organize multiple matrix-like assay datasets. 
However, there are several key distinctions that make `FlexAssays` more suitable 
for certain use cases:

* Flexibility on both rows and columns
* Consistency with matrix-like objects

We will discuss their differences in details below. The `miniACC` dataset from 
`MultiAssayExperiment` will be used for demonstration:

```{r}
library(MultiAssayExperiment)
library(FlexAssays)

data("miniACC")
miniACC
```

We then extract the first 4 datasets from `miniACC` and set the `primary` names 
as their column names, which represent shared samples. A `FlexAssays` is then 
constructed.

```{r}
exps <- experiments(miniACC)[1:4]
for (i in names(exps)) {
  smap <- sampleMap(miniACC)[sampleMap(miniACC)$assay == i, ]
  colnames(exps[[i]]) <- smap$primary[match(colnames(exps[[i]]), smap$colname)]
}
fa <- FlexAssays(exps)
fa
```

As we can see, a `FlexAssays` can hold any matrix-like objects which contain row / column 
names as character vectors and can be subset with `x[i, j]`, including the commonly used 
S4 object `SummarizedExperiment`.

# Extract an assay

For both `MultiAssayExperiment` and `FlexAssays`, an assay can be extracted with 
`[[`:
```{r}
fa[[2]]
```

```{r}
miniACC[[2]]
```

That means both classes can use common extraction methods like a normal `list`. 
The main specification for `FlexAssays` is to use operations supported by a normal
matrix.

# `rownames()` and `colnames()`

For a `MultiAssayExperiment`, `rownames()`/`colnames()` will return a list containing 
corresponding dimension names for each experiment:
```{r}
rownames(miniACC)
```

In contrast, `FlexAssays` returns a character vector of the shared row or column 
names across all assays, just like a standard matrix-like object.

```{r}
rownames(fa)
```

# Subsetting by rows or columns

## Character indices
Since we can treat a `FlexAssays` as a matrix with different layers. The `[i, j]` operation 
should also work well. For example, we subset `fa` with dimension names:

```{r}
i <- sample(rownames(fa), 10)
j <- sample(colnames(fa), 10)
fa[i, j]
```

A `MultiAssayExperiment` can also be subset according to shared dimension names:
```{r}
miniACC[i, j]
```

## Integer indices

A `FlexAssays` can be subset with row / column indices. For example subset the first 
10 shared dimensions:
```{r}
fa[1:10, 1:10]
```

The similar behaviors in `MultiAssayExperiment` are a bit complex.

```{r}
mini2 <- miniACC[1:10, 1:10]
mini2
```

Here the first 10 shared columns can be guaranteed subset.

```{r}
all(rownames(colData(mini2)) == rownames(colData(miniACC))[1:10])
```

However, `[i, ]` simply subset the first 10 rows for each experiment. The extracted 
rows are quite non-predictable.

```{r}
unique(unlist(rownames(mini2)))
```

## logical indices

Subsetting a `FlexAssays` with logical indices is also on the shared rows / columns.

```{r}
i <- sample(c(TRUE, FALSE), nrow(fa), replace = T)
j <- sample(c(TRUE, FALSE), ncol(fa), replace = T)
fa[i, j]
```

For a `MultiAssayExperiment`, the behavior is not clear and may cause an out-of-bounds error:
```{r, error=TRUE}
miniACC[i, j]
```

## Summary

In summary, `FlexAssays` is designed for scenarios where assays differ in both row and column 
dimensions but still benefit from consistent matrix-like behavior. Compared to 
`MultiAssayExperiment`, it provides:

* Familiar and predictable subsetting operations
* Lightweight and consistent behavior with standard matrix classes

These features make it a complementary and user-friendly alternative to `MultiAssayExperiment`, 
especially for use cases in single-cell and spatial data integration workflows.

# Session information

```{r}
sessionInfo()
```
